---
title: "Basic Trends in AISH Trends"
author: 
  - "Andriy Koval"
  - "Kyle Belanger"  
output:
  html_document:
    keep_md: yes
    toc: yes
    toc_float: yes
    code_folding: show
    theme: simplex
    highlight: tango
editor_options: 
  chunk_output_type: console
---

This report explores data on Assured Income for the Severely Handicapped (AISH) taken from the [open.alberta.ca](https://open.alberta.ca/opendata/assured-income-for-the-severely-handicapped-aish-caseload-alberta) portal.

<!--  Set the working directory to the repository's base directory; this assumes the report is nested inside of two directories.-->
```{r, echo=F, message=F, results = 'hide'}
```


```{r set_options, echo=F}
# cat("Working directory: ", getwd()) # turn on to test the location
report_render_start_time <- Sys.time()
# set options shared by all chunks
knitr::opts_chunk$set(
  results      = 'show',
  message      = FALSE,
  warning      = FALSE,
  comment      = NA,
  tidy         = FALSE,
  dpi        = 400, # dots per inch,
  out.width  = "900px", # pixels, this affects only the markdown, not the underlying png file.  The height will be scaled appropriately.
  fig.width    = 11, # inches
  fig.height   = 6, # inches
  out.width    = 1200,
  fig.path     = 'figure-png-com/' # where figures are stored
)
echo_chunks    <- FALSE #Toggle for debugging.
message_chunks <- FALSE #Toggle for debugging.
options(width=100) # number of characters to display in the output (dflt = 80)
ggplot2::theme_set(ggplot2::theme_bw()) # common theme for all graphs
```


<!-- Load the sources.  Suppress the output when loading sources. --> 
```{r load-sources, echo = F}
source("../../scripts/common-functions.R")
```


<!-- Load packages, or at least verify they're available on the local machine.  Suppress the output when loading packages. -->
```{r load-packages, message=message_chunks, echo =F}
library(magrittr)  # pipes
library(dplyr)     # data wrangling
library(ggplot2)   # graphs
library(janitor)   # tidy data
library(tidyr)     # data wrangling
library(forcats)   # factors
library(stringr)   # strings
library(lubridate) # dates
```


<!-- Load any Global functions and variables declared in the R file.  Suppress the output. --> 
```{r declare-globals, echo =F}
# custom function for HTML tables
neat <- function(x, output_format = "html"){ 
  # knitr.table.format = output_format
  if(output_format == "pandoc"){
    x_t <- knitr::kable(x, format = "pandoc")
  }else{
    x_t <- x %>%
      # x %>%
      # neat() %>%
      knitr::kable(format=output_format) %>%
      kableExtra::kable_styling(
        bootstrap_options = c("striped", "hover", "condensed","responsive"),
        # bootstrap_options = c( "condensed"),
        full_width = F,
        position = "left"
      )
  } 
  return(x_t)
}
# Note: when printing to Word or PDF use `neat(output_format =  "pandoc")`

# functionalize
draw_trends <- function(
  d
  ,selected_measure
){
   target_date <- as.Date("2020-04-15")
  d1 <- d %>%
    filter(measure_type == selected_measure)
  g1 <- d1 %>%
    {
    ggplot(., aes(x=date, y=value, color = measure, group = measure))+
    geom_line(size=4, alpha = .1, color = "black" )+
    geom_line(size=.5)+
    geom_point(aes(x=target_date), shape = 21, fill = NA,size = 2, color = "black",
               data = . %>% filter(date == target_date))+
    # geom_vline(xintercept = as.Date("2020-03-11"), linetype="dashed")+
    facet_wrap(vars(!!rlang::sym("measure")), scales = "free_y")+
    scale_y_continuous(labels = scales::comma)+
    labs(
      title = paste0("AISH Caseload by ", selected_measure)
      ,color = "Metric"
      ,x = "Date"
      ,y = "Number of cases"
      ,caption = "Temporal resolution: 1 month       Hollow circle = April 2020"
    )
    }
  g1
}
# Test and Dev
# ds1 %>% draw_trends("Family Composition") 

```


```{r load-data, results='show', message=FALSE, echo =F}
# getwd()
path_file <- "../../data-public/raw/open-alberta/2021-04/goasharedcssspqpqara_research-strategiesopen-data00-open-data-asset-packages-2021aish-monthlyais.csv"
ds0 <- readr::read_csv(path_file) %>% janitor::clean_names()

```

```{r tweak-data,echo = echo_chunks, results='show', message=message_chunks, echo =F}
ds1 <- ds0 %>%
  mutate(
    date = parse_date_time(ref_date,"Y%m%") %>% as_date() + 14 # middle of the month
  )
```

# Take away points 

## Observations

- Caseload experiences a stunted growth after the onset of COVID ( ~ March of 2020) (see [Gender](#gender)). This may be explained by the introduction of the $2,000 monthly assistance to the unemployed due to COVID in April of 2020.  

- This explanation is corroborated in the caseload trends by [Medical Condition](#medical-condition): after April 2020 clients with Physical Disability exhibited a more pronounced plateau than those with Cognitive or Mental Illness Disorders.  

- This trend (stunted growth of caseloads after April 2020) is more pronounced in South and Central regions (see [Regions](#regions)).  

- Employment Participation ([Income](#income)) drops precipitously after April 2020, but recovers to December 2019 levels by the end of 2021.  


## Potential Issues

Some of the observed trends are too drastic to invite an intuitive explanation and therefore are suspect of a data entry issue

- North Central and North West exhibit (see [Regions](#regions)) unnatural change from May to June of 2020. It's as if the data entry was swapped for them starting with June 2020.   

- There is an aberration in caseload among Childless Couples and Couples with Children (see [Family Composition](#family-composition)) for the month of April 2020. These are so uncharacteristic of the preceding and following trajectory that we suspect a data entry.   

- The same month, April 2020 stands out when caseload is broken by [Client Average Age](#age). Clients of Age 65+ exhibit a sharp spike in case counts, while clients of Age 18 - 19 show a sharp depression for that month. Other age groups (e.g. 20-24, 45-49, 50-54) also exhibit unnatural patterns for the month of April 2020 that deviate sharply from previous and following trajectories. 



# Gender
```{r graph-6, echo = echo_chunks, results='show', message=message_chunks, fig.height = 4}
ds1 %>% draw_trends("AISH Client Gender")
```

# Medical Condition
```{r graph-3, echo = echo_chunks, results='show', message=message_chunks}
ds1 %>% draw_trends("AISH Client Primary Medical Condition") 
```

# Regions

```{r graph-7, echo = echo_chunks, results='show', message=message_chunks, fig.height = 7}
ds1 %>% draw_trends("CSS Regions")
```

- __DATA ISSUE__: North Central and North West exhibit (see [Regions](#regions)) unnatural change from May to June of 2020. It's as if the data entry was swapped for them starting with June 2020.   



# Family Composition
```{r graph-1, echo = echo_chunks, results='show', message=message_chunks}
ds1 %>% draw_trends("Family Composition")
```

- __DATA ISSUE__: There is an aberration in caseload among Childless Couples and Couples with Children (see [Family Composition](#family-composition)) for the month of April 2020. These are so uncharacteristic of the preceding and following trajectory that we suspect a data entry error.  

# Caseload
```{r graph-2, echo = echo_chunks, results='show', message=message_chunks}
ds1 %>% draw_trends("AISH Client Caseload")
```



# Age
```{r graph-4, echo = echo_chunks, results='show', message=message_chunks}
ds1 %>% draw_trends("AISH Client Average Age") 
```

- __DATA ISSUE__:  April 2020 stands out when caseload is broken by [Client Average Age](#age). Clients of Age 65+ exhibit a sharp spike in case counts, while clients of Age 18 - 19 show a sharp depression for that month. Other age groups (e.g. 20-24, 45-49, 50-54) also exhibit unnatural patterns for the month of April 2020 that deviate sharply from previous and following trajectories.  

# Income
```{r graph-5, echo = echo_chunks, results='show', message=message_chunks, fig.height = 4}
ds1 %>% draw_trends("AISH Client Income Type Snapshots")
```







Session Information {#session-info}
===========================================================================

For the sake of documentation and reproducibility, the current report was rendered in the following environment.  Click the line below to expand.

<details>
  <summary>Environment <span class="glyphicon glyphicon-plus-sign"></span></summary>
```{r session-info, echo=FALSE}
if( requireNamespace("devtools", quietly = TRUE) ) {
  devtools::session_info()
} else {
  sessionInfo()
}
```
</details>

```{r session-duration, echo=FALSE}
report_render_duration_in_seconds <- round(as.numeric(difftime(Sys.time(), report_render_start_time, units="secs")))
```

Report rendered by `r Sys.info()["user"]` at `r strftime(Sys.time(), "%Y-%m-%d, %H:%M %z")` in `r report_render_duration_in_seconds` seconds.
